---
title: Group by and Mutate Fun
author: Amy Linehan
authors: ["alinehan"]
date: '2020-04-03'
slug: group-by-and-mutate-fun
categories: ["R"]
tags: ["R"]
subtitle: ''
summary: ''
lastmod: '2020-04-03T13:06:22-07:00'
featured: yes
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

I wanted to write this post simply because I use this trick a lot and it's not super complicated but for new programmers, this is often one of the first challenging tasks that can be made simple with dplyr and data.table.

In this post, we're going to use the following data and focus on ranking the ages of each student based on their location. 

```{r include =F}
library(data.table)
library(dplyr) 
```
```{r}

data<-data.table(Location=c("Los Angeles","San Francisco","Los Angeles","New York","New York","New York","Los Angeles","New York","New York","Los Angeles"),
                 Age=c(13,24,13,24,25,18,17,4,29,13))


print(data)

```


So first, let's do this in dplyr:

```{r}

data %>% 
  as.data.frame() %>% 
  group_by(Location) %>% 
  mutate(Location_Rank = 1:n()) %>% 
  arrange(Location, Location_Rank)


```

This isn't revolutionary but for the first couple weeks of consistently using the group by and summarize functions together to analysis/play around with data, it seemed crazy that you could use other functions with group by. So when I discovered that you could use mutate with group by it was really a gamer changer.

As is common in my R blog posts, you can also do this in data.table!!!: 


```{r}
data <- as.data.table(data)

data[, Location_Rank := seq_len(.N), by = Location]
data[, Location_Rank := rowid(Location)]

```



*** introduce example that has you flag and then cumsum 
or look above or below **********



