---
title: Dplyr Functionality with Data.Table
author: Amy Linehan
authors: ["alinehan"]
date: '2020-04-10'
slug: dplyr-functionality-with-data-table
categories:
  - R
tags:
  - R
  - dplyr
  - data.table
subtitle: ''
summary: 'Using data.table to perform common dplyr data cleaning and analyses'
lastmod: '2020-04-10T13:38:07-07:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

For the first year of my career as a consultant, my manager wrote his scripts in data.table and I somehow avoided ever having to really understand what the heck he was doing. It took a full year, but eventually I started understanding and using data.table in my scripts. Now, don't get me wrong I'm definitely still a tidyverse girl and prefer ctrl+alt+m to brackets any day. But, I can't deny that there are situations where using data table is faster, easier and cleaner.

The goal of this post is to consolidate all the common dplyr code I rely on regularly and the corresponding translation in data.table. We'll do another post that dives more into data.table and the pros/cons/tricks to it, but this is more of just a translator between popular dplyr functionalities and data.table. For this, I'll just be using the standard mtcars dataset. 

1. Mutate (add a new flag indicating if mpg is over 20)

```{r}
require(dplyr)
require(data.table)
cars <- mtcars %>% tibble::rownames_to_column(var="car_type")

#dplyr
data_dplyr <- as.data.frame(cars) %>% 
                  mutate(eco_flg = mpg > 20 )
  # data.table 
  data_dt <- as.data.table(cars)
data_dt[,eco_flg:=F]
data_dt <- data_dt[mpg>20,eco_flg:=T]

```

# 2. Summarise and count by a field 

```{r}

# dplyr
data_dplyr %>% group_by(eco_flg) %>% count()

# data.table 
data_dt[,.N,by=eco_flg]

```

# 3. Group by & summarise the average 
```{r}

# dplyr
data_dplyr %>% group_by(eco_flg, gear) %>% summarise(avg_hp = mean(hp))

# data.table 
data_dt[,list(avg_hp= mean(hp)),by=list(eco_flg,gear)]

```


# 4. Join data 

In this example, let's build some sample data to join onto mtcars. Let's say we want to pull data for our friends' cars: 

```{r}

friends <- c("Nellie", "Drake", "JP", "Lily")
cars <- c("Ferrari Dino", "Honda Civic", "Fiat X1-9", "Smart Car")

friends_cars <- data.frame(friends,cars) %>%
  mutate_at(vars(friends, cars), 
            funs(as.character))

head(friends_cars)

```

First, let's join mtcars onto friends using dplyr: 

```{r}

friends_cars %>% left_join(data_dplyr, by = c("cars"= "car_type"))

```

Now, let's try the same join with data.table: 

```{r}

friends_cars <- as.data.table(friends_cars)

friends_cars[]

```
